---
title: "Sampling Plans to Meet Regulatory Requirements for Material Verification"
author: 
  - name: Joel Anderson
    email: janderson@ffs-integrity.com
    affiliation:
      - name: Total Integrity Analytics
format: 
  docx:
   toc: true
   number-sections: true
editor: visual
date: today
date-format: long
project:
  output-dir: C:/Temp/quarto-output/sampling
execute:
  echo: false
  warning: false
  message: false
  dpi: 300
---

# Executive Summary

This paper provides a practical framework for meeting the material verification requirements of 49 CFR 192.607 without relying solely on prescriptive rules. It explains how to judge whether a sample is inconsistent with a reference population and how to achieve the 95% confidence level required for an expanded sampling program.

Current regulations require operators to verify pipeline material properties when traceable, verifiable, and complete (TVC) records are missing. While the default approach mandates one excavation per mile—up to 150 digs—this can be costly and inefficient.

This document introduces statistically valid sampling methods that achieve the required 95% confidence level with far fewer samples. Key principles include:

-   **Risk-Based Decision Making**: Define acceptable error thresholds based on the consequences of Type I (false positive) and Type II (false negative) errors. These thresholds drive sample size and test design.
-   **Measurement Integrity**: Account for uncertainty in measurements and material variability. Use confidence intervals—not just sample means—to avoid misclassifying materials and triggering unnecessary expanded sampling programs.
-   **Optimized Sampling via Bayesian Logic**: Apply Bayes’ Theorem to incorporate prior knowledge (e.g., manufacturing standards, operating conditions) and update probabilities as new data are collected. This approach can reduce required excavations to far fewer than 150.
-   **Expanded Sampling**: Determine the sample size needed to achieve the 95% confidence level required under 192.607(e)(5). Two approaches are presented: a traditional Frequentist method and a Bayesian method.

By combining Frequentist and Bayesian methods, operators can design sampling plans that are regulatory compliant, statistically defensible, and economically sustainable. These strategies enable operators to demonstrate material integrity with confidence while minimizing cost and operational disruption.

# Regulatory Requirements and Context

This section summarizes the key requirements in 49 CFR 192.607 that drive sampling design. The sections that follow move from statistical foundations (why sample sizes behave the way they do) to practical methods (how to set decision thresholds, calculate sample sizes, and design expanded/alternate sampling programs).

```{r}
#| label: setup

library(tidyverse)
library(patchwork)
library(tidymodels)
library(pwr)
library(infer)
library(viridis)
library(scales)
library(here)
library(flextable)
library(rriskDistributions)

set.seed(42)

theme_sampling <- function(base_size = 12, base_family = "sans") {
    theme_minimal(base_size = base_size, base_family = base_family) +
        theme(
            # Typography
            text = element_text(family = base_family, color = "#2c3e50"),
            plot.title = element_text(
                size = base_size + 4,
                face = "bold",
                color = "#2c3e50",
                margin = margin(b = 10)
            ),
            plot.subtitle = element_text(
                size = base_size + 1,
                color = "#7f8c8d",
                margin = margin(b = 15)
            ),
            plot.caption = element_text(
                size = base_size - 2,
                color = "#95a5a6",
                hjust = 0
            ),
            axis.title = element_text(
                size = base_size + 1,
                face = "bold",
                color = "#2c3e50"
            ),
            axis.text = element_text(
                size = base_size,
                color = "#34495e"
            ),

            # Grid and background
            panel.grid.major = element_line(color = "#ecf0f1", linewidth = 0.5),
            panel.grid.minor = element_line(color = "#f8f9fa", linewidth = 0.25),
            panel.background = element_rect(fill = "white", color = NA),
            plot.background = element_rect(fill = "white", color = NA),

            # Legend
            legend.title = element_text(face = "bold", color = "#2c3e50"),
            legend.text = element_text(color = "#34495e"),
            legend.background = element_rect(fill = "white", color = "#ecf0f1"),
            legend.box.background = element_rect(fill = "white", color = "#ecf0f1"),

            # Facets
            strip.background = element_rect(fill = "#3498db", color = "#3498db"),
            strip.text = element_text(color = "white", face = "bold"),

            # Margins and spacing
            plot.margin = margin(20, 20, 20, 20),
            panel.spacing = unit(1, "lines")
        )
}

# Modern color palette
sampling_colors <- c(
    "primary" = "#3498db", # Blue
    "secondary" = "#e74c3c", # Red
    "success" = "#27ae60", # Green
    "warning" = "#f39c12", # Orange
    "info" = "#9b59b6", # Purple
    "light" = "#ecf0f1", # Light gray
    "dark" = "#2c3e50" # Dark gray
)

```

The regulations published by PHMSA in October 2019 require non-destructive examination (NDE) to determine the material properties of pipelines that lack traceable, verifiable, and complete (TVC) records. To meet this requirement, operators must verify pipeline material properties in accordance with 49 CFR 192.607.

Section 192.607(e)(1) states that a population is defined by a combination of material properties and attributes, including:

-   Nominal wall thickness
-   Grade
-   Manufacturing process
-   Pipe manufacturing dates
-   Construction dates

Pipeline segments within a population do not need to be contiguous.

If an operator follows the prescriptive requirements in 192.607(e)(2), they must perform one excavation per mile for each population (rounded up to the nearest whole number), up to a maximum of 150 excavations when the population exceeds 150 miles.

If an operator uses an alternate statistical sampling approach, 192.607(e)(5) allows another method with a statistically valid basis designed to achieve a 95% confidence level.

The goal of a sampling program is to sample enough to be able to make a defensible decision about the population without sampling too much. Too small of a sample increases the risk of making an erroneous decision about the population, sampling too much wastes time and money, indirectly increasing risk elsewhere by diverting resources unnecessarily.

Perfect measurements are impossible, so differences between measured and true values will always exist. Errors of varying magnitude will exist due to random effects from any number of sources. The key question: Are the differences between the sample and the hypothesized value simply random variation, or do they indicate a real discrepancy? Additionally, have enough samples been completed to be able to tell the difference to a level of precision that minimizes the risk to an acceptable level while also minimizing the cost and complexity.

There is no universal number that is applicable to all situations. The necessary sample size is interrelated with the acceptable risk. Changes to the acceptable risk will change the number of samples significantly. With some knowledge of sampling statistics, the risks can be lowered to an acceptable level. This is the fundamental premise of statistical sampling: the process of selecting a representative subset from a larger population to estimate the properties of the whole.

# Statistical Foundations

The goals of sampling fall into two general categories. The first is to compare a sample to some reference value or two samples against each other. In this type of sampling the measurements can take on any number on a continuous scale. The second type of sampling, the interest is only whether the sample meets (or not) some criterion analogous to quality control. In this case, the data collected is the number that met the criterion out of the total number sampled. The metric of this type of sampling is an inferred proportion of the population that meets the requirements.

## Lot Size vs. Sample Size Myth

As an introduction to sampling, a few sampling principles will be introduced to dispel some entrenched myths.

**Principle I:** The size of the population (the "lot" in quality control terms) does not appreciably change the information obtained from the sample. Assuming the lot is much greater than the sample size by at least a factor of 10. This means that the sample will contain roughly the same defective percentage regardless of the lot size.

This principle is often the least understood because standards like ANSI/ASQ Standard Z1.4 contain sampling charts that require differing sample sizes depending on the size of the lot. The reason for the sample size to lot size dependency in these standards lies not in statistics but economics. As the lot size grows the economic benefit of larger samples is worth it to acquire more information about it. This leads to Principle II.

**Principle II:** The information contained in a sample increases when the size of the sample increases, independent of the size of the lot. The larger the sample, the more faithful the sample will be to the population.

**Principle III:** The information contained in a sample does not depend on the *proportion* that the sample represents of the lot, only the overall size. If we sample 100 from a lot of 10,000 it will more closely represent the population than a sample of 10 from a population of 100 even though the sample of 10 is a much larger proportion of the population.

Principle I is demonstrated in @fig-lotsize where 5,000 random samples of size 10 were taken from two different size populations. One has a size of 100 and the second is 10,000. Both lots have a 10% defect rate. For each sample taken, the total number of defective items is collected. The results of the two different lot sizes is nearly identical to the theoretical binomial theorem results. This also shows the variability of sampling; in all three cases the defect rate for the populations are fixed at 10% but the number of defective ranged from 0 to 6 out of 10 sampled. This wide variability is partially due to the relatively small sample size. Increasing the sample size from 10 to larger number would decrease the variance in accordance with Principle II.

```{r}
#| label: fig-lotsize
#| fig-cap: "Size of Lot Doesn't Affect the Outcome"
#| dpi: 300
#| fig-width: 6.5

# =============================================================================
# CONFIGURATION PARAMETERS
# =============================================================================
SAMPLE_SIZE <- 10 # Size of each sample
NUM_SAMPLES <- 5e3 # Number of samples to take
DEFECTIVE_RATE <- 0.10 # 10% defective rate

# =============================================================================
# POPULATION DEFINITION
# =============================================================================
# Create two populations with same defective rate but different sizes
# Principle: Population size doesn't matter, only the percentage defective

lot_large <- tibble(
    defective = c(rep(0, 9e3), rep(1, 1e3)) # 10K items, 10% defective
)

lot_small <- tibble(
    defective = c(rep(0, 90), rep(1, 10)) # 100 items, 10% defective
)

# =============================================================================
# SAMPLING SIMULATION
# =============================================================================
# Take repeated samples from both populations

samples_large <- rep_sample_n(lot_large, size = SAMPLE_SIZE, reps = NUM_SAMPLES)
samples_small <- rep_sample_n(lot_small, size = SAMPLE_SIZE, reps = NUM_SAMPLES)

# Combine samples and calculate defective counts
sampling_results <- samples_large %>%
    bind_cols(defective_small = samples_small$defective) %>%
    summarise(
        defective_large = sum(defective),
        defective_small = sum(defective_small),
        .groups = "drop"
    )


# =============================================================================
# THEORETICAL VS EMPIRICAL COMPARISON
# =============================================================================


# Theoretical binomial distribution
theoretical_counts <- tibble(
    defective_count = 0:5,
    theoretical_prob = dbinom(defective_count, size = SAMPLE_SIZE, prob = DEFECTIVE_RATE),
    expected_frequency = round(theoretical_prob * NUM_SAMPLES, 0)
)


# Empirical distribution
empirical_distribution <- sampling_results %>%
    pivot_longer(
        cols = c(defective_large, defective_small),
        names_to = "population",
        values_to = "defective_count"
    ) %>%
    group_by(population, defective_count) %>%
    summarise(
        empirical_frequency = n(),
        empirical_prob = empirical_frequency / NUM_SAMPLES,
        .groups = "drop"
    ) %>%
    mutate(
        population = case_when(
            population == "defective_large" ~ "Large Population",
            population == "defective_small" ~ "Small Population"
        )
    )


# =============================================================================
# ENHANCED THEORETICAL VS EMPIRICAL PLOT
# =============================================================================

# Combine theoretical and empirical data
comparison_data <- theoretical_counts %>%
    select(defective_count, theoretical_prob) %>%
    mutate(population = "Theoretical") %>%
    bind_rows(
        empirical_distribution %>%
            select(defective_count, empirical_prob, population) %>%
            rename(theoretical_prob = empirical_prob)
    )

# Create enhanced comparison plot
enhanced_comparison <- comparison_data %>%
    ggplot(aes(
        x = defective_count, y = theoretical_prob,
        color = population, shape = population
    )) +
    geom_line(linewidth = 1.1, alpha = 0.8) +
    geom_point(size = 2.5, alpha = 0.8) +
    scale_color_viridis_d(option = "viridis") +
    scale_shape_manual(values = c(16, 17, 18)) +
    theme_sampling() +
    labs(
        title = "Theoretical vs Empirical Distributions",
        subtitle = "Comparison of Probabilities",
        x = "Number of Defective Items",
        y = "Probability",
        color = "Distribution",
        shape = "Distribution",
        caption = "Theoretical: Binomial(n=10, p=0.1) | Empirical: Based on simulation"
    ) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1))

print(enhanced_comparison)


```

The key takeaway from @fig-lotsize is that the size of the population doesn't affect the outcome of a sample. There is no need to take a larger sample simply because the population is larger. This demonstrates that the prescriptive one-excavation-per-mile requirement is not rooted in statistical necessity but in administrative simplicity.

# The Population vs. Sample Paradigm {#sec-populationsample}

## Population vs. Sample

In statistics, the concepts of population and sample form the foundation of inferential analysis. The population refers to the entire group of individuals, objects, or events that you're interested in studying. For example, if you're researching voter preferences in a country, the population might be all eligible voters—potentially millions of people. The population is typically large, sometimes infinite (like all possible rolls of a die), and often impractical or impossible to measure fully due to time, cost, or logistical constraints.

A sample, on the other hand, is a subset of the population selected for actual measurement or observation. It's chosen to represent the population as accurately as possible, but it's always smaller and more manageable. Using the voter example, a sample might be 1,000 randomly selected voters surveyed via phone or online. The goal is to use data from the sample to make inferences about the population—such as estimating the average (mean) preference or the proportion supporting a candidate—without needing to query everyone. This relationship is often called the "population-sample paradigm," where the sample acts as a proxy for the population.

## Sampling Distributions

The Role of Sample Size: Smaller samples tend to vary more because they're more susceptible to outliers or unrepresentative selections. As sample size increases, variability decreases (per the Law of Large Numbers), and sample statistics get closer to population parameters. However, even large samples aren't identical; they just cluster more tightly around the true value. This is shown in @fig-sampling-variability where different sample sizes are taken from the same population distribution. The larger the sample size, the narrower the distribution of sample means, indicating less variability.

```{r}
#| label: fig-sampling-variability
#| fig-cap: "Sampling Variability Relative to Sample Size"
#| fig-width: 6.5
#| fig-height: 4
#| dpi: 300

wtpars <- rriskDistributions::get.beta.par(
  q = c(0.95 * .375, 0.375, 1.1 * 0.375),
  show.output = FALSE,
  plot = FALSE
)

wall <- tibble(wt = rbeta(1e4, wtpars[[1]], wtpars[[2]]))

n1 <- 10
n2 <- 25
n3 <- 50
n4 <- 100
nzz <- c(n1, n2, n3, n4)

nsamp <- 1e4

# Bootstrap samples -------------------------------------------------------

# bootstrapping samples of various sizes, if I was to take samples of size "x"
# multiple times from this population what would it look like? This demos the
# likely spread and variability in the mean based on sample size

map(nzz, ~ rep_slice_sample(
  .data = wall,
  reps = nsamp,
  replace = TRUE,
  n = .
)) %>%
  tibble(s = .) %>%
  mutate(n = nzz) %>%
  unnest(cols = s) |> 
  mutate(n = factor(n)) %>%
  group_by(replicate, n) %>%
  summarise(mwt = mean(wt)) |>
  ggplot(aes(mwt, col = factor(n))) +
  geom_density(adjust = 5, lwd = 1, n = 501) +
  theme_minimal(12, "serif") +
  labs(
    title = "Density Curves Based on Sample Size",
    col = "Sample Size",
    y = NULL,
    x = "Mean of Sample"
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(0.7, 0.7, 0.7, 0.9, "cm")
  )

```

Sampling Distribution Concept: If you repeatedly draw samples of the same size from the population and calculate a statistic (like the mean) for each, the distribution of those statistics is called the sampling distribution. For means, the Central Limit Theorem tells us this distribution is approximately normal for large samples, with its own mean equal to the population mean and standard deviation (standard error) of σ/√n (where n is sample size). The spread of this distribution explains why samples vary: most sample means fall within a few standard errors of the population mean, but outliers can occur. As an example of this, 5,000 samples, each with a size of 10 are drawn from a population representing male heights in the United States (mean = 69 inches, standard deviation = 3.0). The distribution of the sample means is shown in @fig-sampling-distribution.

```{r}
#| label: fig-sampling-distribution
#| fig-cap: "Sampling Distribution of the Mean from 5,000 Samples of Size 10"
#| fig-width: 6
#| fig-height: 4

samp_size <- 10 # size of sample
nsamps <- 5000 # number of samples we will take

# Define population parameters
pop_mean <- 69
pop_sd <- 3.0

adult <- tibble(height = rnorm(1e5, pop_mean, pop_sd))

# Optimized sampling: Sample directly from vector and compute mean immediately
# This avoids creating/unnesting a 50,000 row dataframe
boots <- tibble(
    mh = replicate(nsamps, mean(sample(adult$height, samp_size)))
)


# plot results ------------------------------------------------------------
# Now let's plot the sampling distribution. We will also overlay the sampling
# distribution of the mean predicted on the basis of the population mean and
# standard deviation, to show that it properly describes the actual sampling
# distribution.

# pipe the boots data frame into ggplot
boots |>
    ggplot(aes(mh)) +
    # create histogram using density rather than count
    geom_histogram(
        aes(y = after_stat(density)),
        bins = 50,
        col = "white",
        fill = "steelblue",
        alpha = 0.7
    ) +
    # add a vertical line for the population mean
    geom_vline(
        xintercept = pop_mean,
        linewidth = 1,
        color = "firebrick",
        linetype = "dashed"
    ) +
    # add a label for the line
    annotate(
        "text",
        x = pop_mean,
        y = dnorm(
            pop_mean, pop_mean,
            pop_sd / sqrt(samp_size)
        ),
        label = "Population Mean",
        size = 4,
        hjust = -0.2,
        color = "firebrick",
        fontface = "bold"
    ) +
    # label the x axis
    labs(
        title = "Sampling Distribution of the Mean",
        subtitle = paste("Sample Size =", samp_size),
        x = "Height (inches)",
        y = "Density"
    ) +
    # add normal based on population mean/sd
    stat_function(
        fun = dnorm,
        n = 200,
        args = list(
            mean = pop_mean,
            sd = pop_sd / sqrt(samp_size)
        ),
        linewidth = 1,
        color = "black",
        linetype = "dashed"
    ) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    theme_minimal(base_size = 12)
```

The key takeaway from @fig-sampling-distribution is that even though all samples were drawn from the same population using the same method, the sample means vary around the true population mean. This variability is inherent to any sampling process and underscores the importance of understanding sampling distributions when making inferences about populations based on sample data and why we need hypothesis testing to make decisions based on sample data.

## Hypothesis Testing {#sec-hypothesis-testing}

A hypothesis test is a statistical method used to decide if the data from a sample provides enough evidence to support a claim or assumption (the hypothesis) about the larger population.

Most elementary statistics are introduced with concept an urn with a known number of balls itemized by color. Everything about the contents of the urn is known and the goal is to predict what kind of data it will produce. While drawing balls from an urn is a good analogy to introduce statistics, it is not representative of any problem of consequence. In almost any problem outside of contrived examples and casino games, neither the size nor the exact make-up of the population is known ahead of time. Rather than predicting what data will come out of drawing from the urn, the engineer has a much different problem, that is to reason about the likely contents of the urn based on their existing knowledge of the problem and a limited set of observations. Therefore, any statistic should not be confused with the reality of nature. A statistic is a measure of our state of knowledge about the problem at that time. The true contents of the urn remain an unknown because in most situations, sampling the entire population is either impractical or impossible.

In any hypothesis test, there are only two possible outcomes: either reject the hypothesis because the observed data is highly improbable if that hypothesis were true, or fail to reject it if the data remains consistent with the premise. We can never 'accept' a hypothesis as absolute truth, as doing so would require us to rule out an infinite number of alternative possibilities. For example, if we hypothesize that the mean yield strength is 42 ksi, accepting that value would require us to prove it isn't 42.1, 42.01, or any other value across infinite levels of precision.

An hypothesis test is structurally analogous to a criminal trial. The defendant is assumed innocent ($H_0$) and is ultimately pronounced either 'guilty' or 'not guilty'—never 'innocent'. Just as a trial seeks to prove guilt beyond a reasonable doubt rather than all doubt, a statistical test measures the weight of evidence against a baseline.

Understanding this logic is essential for navigating the two paths to regulatory compliance provided in this paper. The first path uses Frequentist statistics (detailed in @sec-measurement-uncertainty), which focuses on comparing current sample data against historical records or reference values to ensure material consistency. The second path utilizes Bayesian logic (detailed in @sec-alternate-sampling-plan), which is the most efficient method for material verification when records are missing. While the Frequentist approach relies on the 'weight of evidence' from a single set of data, the Bayesian approach allows us to 'update our beliefs' sequentially, using each successful test to reach the required 95% confidence level with significantly fewer excavations than the prescriptive 150-dig maximum. By defining our risks and hypotheses upfront, we can choose the method that best balances safety with operational efficiency.

## Types of Errors {#sec-statistical-tests-and-errors}

Before moving on to determining the necessary sample size, it is helpful to clarify statistical errors.

Any statistical test starts with the null hypothesis ($H_0$): the assumption that there is no difference between groups (or between a sample and a reference value). The test framework is commonly referred to as a Null Hypothesis Significance Test (NHST).

In NHST, there are two types of errors:

-   **Type I error** ($\alpha$): Rejecting $H_0$ when $H_0$ is true (a **false positive**).
-   **Type II error** ($\beta$): Failing to reject $H_0$ when $H_0$ is false (a **false negative**).

The plot in @fig-type1and2 shows Type I and Type II errors graphically. Here, $H_0$ is the distribution under the null hypothesis and $H_A$ is the distribution under the alternative hypothesis. The darker shaded regions correspond to the error probabilities $\alpha$ and $\beta$. The boundary between the shaded and unshaded areas represents the decision threshold: values beyond that threshold lead to rejection of $H_0$.

These errors exist because, even at a reasonable decision threshold, there is always overlap between the distributions. As a result, there is always some probability of making the wrong decision.

```{r}
#| label: fig-type1and2
#| fig-cap: "Type I and II errors"
#| fig-width: 6.5
#| fig-height: 4
#| dpi: 300

label1 <- c("H[0]", "H[A]")
DF <- data.frame(x = seq(-3, 5, length.out = 301))

typeI <- ggplot(DF, aes(x)) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "steelblue2",
        alpha = .3,
        n = 301
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "steelblue2",
        alpha = 0.85,
        xlim = c(qnorm(0.95), 4)
    ) +
    stat_function(
        fun = dnorm,
        geom = "line",
        linetype = 2,
        col = "black",
        alpha = .5,
        args = list(
            mean = 2
        ),
        n = 301
    ) +
    labs(
        title = "Type I Error",
        x = "Test Statistic",
        y = NULL,
        caption = "Type I = Reject Null When it is True"
    ) +
    scale_x_continuous(limits = c(-3, 5)) +
    annotate(
        geom = "text",
        x = 0,
        y = 0.3,
        label = deparse(bquote(H[0])),
        parse = T,
        size = 5
    ) +
    annotate(
        geom = "text",
        x = 2,
        y = 0.3,
        label = deparse(bquote(H[A])),
        parse = T,
        size = 5
    ) +
    annotate("text",
        x = 2.0,
        y = 0.025,
        label = expression(alpha)
    ) +
    theme_minimal(13, base_family = "serif") +
    theme(axis.text.y = element_blank())

typeII <- ggplot(DF, aes(x)) +
    stat_function(
        fun = dnorm,
        geom = "line",
        linetype = 2,
        n = 301
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "orange",
        args = list(
            mean = 2
        ),
        xlim = c(-2, qnorm(0.95)),
        n = 301
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        linetype = 2,
        fill = "orange",
        alpha = .3,
        args = list(
            mean = 2
        ),
        n = 301
    ) +
    labs(
        title = "Type II Error",
        x = "Test Statisitic",
        y = NULL,
        caption = "Type II = Failure to Reject Null When it is False"
    ) +
    scale_x_continuous(limits = c(-3, 5)) +
    annotate(
        geom = "text",
        x = 2,
        y = 0.3,
        label = deparse(bquote(H[A])),
        parse = T,
        size = 5
    ) +
    annotate(
        geom = "text",
        x = 0,
        y = 0.3,
        label = deparse(bquote(H[0])),
        parse = T,
        size = 5
    ) +
    annotate("text",
        x = 0.5,
        y = 0.03,
        label = expression(beta),
        fontface = "bold"
    ) +
    theme_minimal(13, "serif") +
    theme(axis.text.y = element_blank())

(
    typeI + typeII
) +
    plot_annotation(
        title = "Type I and Type II Error Visualization",
        theme = theme(plot.margin = margin(0.7, 0.7, 0.7, 0.7, "cm"), title = element_text(size = 14))
    )


```

The conclusion from @fig-type1and2 is that Type I and II errors are interrelated. Minimizing the Type I error by setting the threshold for rejection higher (i.e., a lower $\alpha$), increases the likelihood of a Type II error for a given sample size. This begs the question of which error is worse. The answer is, as in most of statistics, it depends. Typically, the acceptable Type I error is set at 0.05 in most tests because it is usually, more consequential. For example if a new drug is being tested, the Null would be that there is no difference in outcome vs. the placebo. In this example, a Type I error would be determining that the new drug has an effect on a disease when it really doesn't and the Type II error would be rejecting that the drug had an effect when it did. The consequences of the Type I far outweigh the Type II in that scenario. But the 0.05 is only a convention, not an absolute.

If a criminal trial were a statistical test, the Null would be that the defendant is innocent (innocent until proven guilty) and the desirable outcome would be that the likelihood of judging an innocent defendant guilty (Type I error) to be as small as possible even at the expense of increasing the likelihood of a guilty defendant going free (Type II error). Setting a low $\alpha$ value means that you are requiring a high level of evidence before the Null is rejected.

```{r}
#| label: tbl-decisions
#| tbl-cap: "Decisions and Errors"

library(ftExtra)

read_csv(here("decisions.csv")) |>
    rename("Real World: H0 is True" = `Real World: H0? is True`) |>
    rename("Real World: H0 is False" = `Real World: H0? is False`) |>
    mutate(
        Decision = str_remove(string = Decision, pattern = "\\?"),
        Decision = c("Reject H~0~", "Fail to Reject H~0~")
    ) |>
    flextable() |>
    compose(part = "header", j = "Real World: H0 is True", value = as_paragraph("Real World: H", as_sub("0"), " is True")) |>
    compose(part = "header", j = "Real World: H0 is False", value = as_paragraph("Real World: H", as_sub("0"), " is False")) |>
    colformat_md() |>
    flextable::autofit()

```

In some circumstances a Type II error has a higher consequence and it is appropriate to set a lower evidence threshold to reject the Null by setting a lower $\beta$ value. In this alternate scenario suppose there are two lots of pipe in a storage yard, one with documentation and another without. It is suspected that the yield strength is the same for the two lots, so a test program is set up to test them and compare the results. If it is judged that the Null (i.e. that they are the same yield strength) can't be rejected, then the undocumented lot will be used along with the documented one for a project. In this case, the Type I error would be rejecting that they are the same when they really are and consequently not using the pipe that lacks documentation, a cost consequence but not a safety one. But a Type II error would mean that the undocumented lot gets used in the project under the assumption that the material properties are the same when they are not. In this case, the consequences of a Type II are much greater than a Type I and therefore a lower acceptable Type II would be desirable. Anytime you want prioritize detection over false alarms you would set $\beta \lt \alpha$.

The ability of a test to correctly reject the Null is known as the power and is equal to $1-\beta$. There are two ways to increase the power of a test, the first is to lower the acceptable Type II error, which in turn raises the Type I error, and the other is to increase the sample size. See @sec-sample-size for a more detailed explanation. In the previous scenario, a higher Type I error was an acceptable trade-off for the benefit of increasing the probability of properly rejecting the Null. If a higher Type I is not acceptable then the alternative is to increase the sample size.

## One-Tail vs. Two-Tail Test

Following the discussion of error types it is necessary to cover the types of tests and how these errors factor into them. The t-test is a statistical tool to determine if the means of two groups (two-sample test) or one group relative to an hypothesized value (one-sample test) are significantly different. A one-tailed t-test is used when the researcher is interested in only one direction of difference between the two groups. For example, if we want to know whether the mean of Group A is significantly greater (or less than) than Group B, we would use a one-tailed t-test. A two-tailed t-test, is used when the researcher is interested if there is any difference between the two groups in either direction. For example, if we want to know whether the mean of Group A is significantly different from Group B, we would use a two-tailed t-test.

In a one-tailed test, the entire $\alpha$ level is in one tail, while in a two-tailed test, the $\alpha$ level is split in half between the two tails. One-tailed tests detect an effect in one specific direction, while two-tailed tests consider effects in both directions. One-tailed tests require prior knowledge about the direction of the effect, whereas two-tailed tests do not. The difference in the two tests are shown in @fig-onetail_twotail, plots A & B are examples of one-tail tests for left and right tails respectively and plot C shows a two-tail test.

```{r}
#| label: fig-onetail_twotail
#| fig-cap: "One-Tail vs. Two-Tail Test"
#| dpi: 300
#| fig-height: 6
#| fig-width: 6.5

DF <- data.frame(x = seq(-3.5, 3.5, length.out = 301))

rtail <- ggplot(DF, aes(x)) +
    stat_function(
        fun = dnorm,
        geom = "area",
        n = 301,
        fill = "royalblue",
        xlim = c(-4, qnorm(0.95)),
        alpha = 0.3
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "orange",
        xlim = c(qnorm(0.95), 4),
        alpha = 0.80
    ) +
    annotate(
        "text",
        x = 2,
        y = 0.025,
        label = expression(alpha),
        size = 4.5
    ) +
    # xlim(-3,3)+
    theme_void(14) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    labs(
        x = NULL,
        y = NULL,
        title = latex2exp::TeX("$H_A:\\; \\mu>\\mu_0$")
    ) +
    scale_x_continuous(breaks = c(0)) +
    geom_vline(lty = 2, xintercept = 0) +
    theme(
        plot.margin = margin(0.5, 0.5, 0.6, 0.5, "cm"),
        plot.title = element_text(hjust = 0.5)
    )


ltail <- ggplot(DF, aes(x)) +
    stat_function(
        fun = dnorm,
        geom = "area",
        n = 301,
        fill = "royalblue",
        xlim = c(qnorm(0.05), 4),
        alpha = 0.3
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "orange",
        xlim = c(-4, qnorm(0.05)),
        alpha = 0.80
    ) +
    annotate(
        "text",
        x = -2,
        y = 0.025,
        label = expression(alpha),
        size = 4.5
    ) +
    # xlim(-3,3)+
    theme_void(14) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    labs(
        x = NULL,
        y = NULL,
        title = latex2exp::TeX("$H_A:\\; \\mu<\\mu_0$")
    ) +
    scale_x_continuous(breaks = c(0)) +
    geom_vline(lty = 2, xintercept = 0) +
    theme(
        plot.margin = margin(0.5, 0.5, 0.6, 0.5, "cm"),
        plot.title = element_text(hjust = 0.5)
    )


twotail <- ggplot(DF, aes(x)) +
    stat_function(
        fun = dnorm,
        geom = "area",
        n = 301,
        fill = "royalblue",
        xlim = c(qnorm(0.025), qnorm(0.975)),
        alpha = 0.3
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "orange",
        xlim = c(-4, qnorm(0.025)),
        alpha = 0.80
    ) +
    stat_function(
        fun = dnorm,
        geom = "area",
        fill = "orange",
        xlim = c(qnorm(0.975), 4),
        alpha = 0.80
    ) +
    annotate(
        "text",
        x = c(-2.15, 2.15),
        y = c(0.02, 0.02),
        label = expression(alpha / 2),
        size = 4
    ) +
    # xlim(-3,3)+
    theme_void(14) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    labs(
        x = NULL,
        y = NULL,
        title = latex2exp::TeX("$H_A:\\; \\mu\\neq\\mu_0$")
    ) +
    scale_x_continuous(breaks = c(0)) +
    geom_vline(lty = 2, xintercept = 0) +
    theme(
        plot.margin = margin(0.5, 0.5, 0.6, 0.5, "cm"),
        plot.title = element_text(hjust = 0.5)
    )

(
    ltail + rtail
) / twotail +
    plot_annotation(
        title = "Left, Right and Two-Tail Test",
        tag_levels = "A",
        tag_suffix = ")"
    )

```

## Power and Effect Size {#sec-sample-size}

The instinctive question in designing a sampling program is, "how much is enough?". The answer to this is dependent on the answer to several other questions. The first one is how big is the effect that the researcher wants to detect. The effect size can be thought of as how much of a difference between the sample and the hypothesized value is a practical concern. The other question that must be answered is what is the acceptable error rate? This ultimately involves defining your acceptable risk, it is impossible to reduce the potential error to zero outside of sampling the entire population, so the person designing the sampling plan must decide what is the acceptable probability of an error relative to the risks involved with an incorrect decision.

There is always going to be some difference between the sample and the true value. Therefore, the researcher needs to decide prior to sampling, how large the difference can be before it becomes a concern. If the testing program is trying to differentiate a mean of 10 from a mean of 10.5, that is a much smaller effect size and requires a much larger sample than trying to differentiate a mean of 10 from a mean of 15. The size of your sample is also dependent on the required power of the test, which is also chosen by the researcher when designing the sampling plan. Recall that the power of a test is simply $1 - \beta$ and that $\beta$ is the Type II error rate as discussed in the section on error types. The larger the sample size, the larger the power of the test (i.e., the smaller the Type II error). The power can be calculated after the data is collected but normally a minimum power is specified prior to sampling in addition to the significance ($\alpha$) which will in turn determine the required sample size.

As a general rule, it is desirable for a test to have enough samples to achieve a minimum power of 80% which corresponds to a 20% Type II error. But as was discussed in @sec-statistical-tests-and-errors, the size of the acceptable $\beta$ is dependent on the relative consequences of Type II vs Type I error. The reason the increased sample size creates increased power of the test is that as the sample size grows, the variability in the mean shrinks, making it easier to differentiate between two hypotheses.

Note that the uncertainty in the mean is not linear with sample size, it's inversely proportional to, $\sqrt{n}$ therefore, to shrink the uncertainty by half, it requires a sample four times as large. The power versus sample size for various effect sizes and the number of samples required to reach 80% power is shown in @fig-sample_size_plots. The effect size (d) is calculated by taking the difference of the two means and dividing by the pooled standard deviation as shown in @eq-cohens. The pooled standard deviation calculation is shown in @eq-ttest.

@fig-sample_size_plots shows the accumulation of power vs. effect size for different sample sizes. The larger samples can acheive 80% power for smaller effect sizes. Whereas the smaller sample sizes would need extremely large differences in the two samples before they could reach 80%. For example to detect a "large" effect size (0.8) your sample size would have to be about 25 to achieve 80% power. Notice how the curves take on a sigmoidal S-shape where the slope is steeper through the middle relative to the ends. This shows that small changes in the effect size in this region can prouce large changes in power. Whereas if the power is near 90% the curve assyptotes and changes in effect size have minimal changes in power. In summary, the required sample size is dependent on the size of effect the researcher wants to detect and the acceptable Type I and II error rate. Power analysis does not replace effect sizes or confidence intervals — it's a planning tool within (or complementary to) NHST.

$$
d = \frac{\bar{x_1}-\bar{x_2}}{S_p}
$$ {#eq-cohens}

```{r}
#| label: fig-sample_size_plots
#| fig-cap: "Sample Size vs. Power for Two Sample t-test"
#| fig-width: 6
#| fig-height: 4
#| dpi: 300


# Function to calculate power for two-sample t-test
calculate_power <- function(n, effect_size, alpha = 0.05, alternative = "two.sided") {
  power <- power.t.test(
    n = n,
    delta = effect_size,
    sd = 1, # Standardized effect size (Cohen's d)
    sig.level = alpha,
    type = "two.sample",
    alternative = alternative
  )$power
  return(power)
}

# Define parameters
sample_sizes <- c(10, 25, 50, 100)
effect_sizes <- seq(0, 1.5, length.out = 301)
alpha <- 0.05

# Create a data frame with all combinations
power_data <- expand.grid(
  n = sample_sizes,
  effect_size = effect_sizes
) %>%
  rowwise() %>%
  mutate(power = calculate_power(n, effect_size, alpha)) %>%
  ungroup() %>%
  mutate(n_label = paste("n =", n))

# Create the plot
ggplot(power_data, aes(x = effect_size, y = power, color = factor(n), group = n)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red", alpha = 0.7) +
  # scale_color_viridis_d(name = "Sample Size", option = "plasma") +
  labs(
    title = "Statistical Power vs Effect Size",
    subtitle = paste("Two-sample t-test, α =", alpha, "(dashed line shows 80% power)"),
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power",
    caption = "Assumes equal group sizes and standardized effect sizes",
    color = "Sample\nSize"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90")
  ) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 1.5, 0.25))

```

The takeaway from @fig-sample_size_plots is that the effect size makes a significant difference on the size of the sample required to reach the necessary power for a sampling plan. While more data points is always better, the cost of collecting additional data is a real world concern and should be weighed against the benefit of further reducing the likelihood of a Type II error relative to the current sample size.

# Measurement Uncertainty {#sec-measurement-uncertainty}

## Measurement Uncertainty and A Single Sample

A measurement is the process of assigning a quantity to a physical property. Regardless of the precision, there is some level of error. The term, "error" in this context does not imply a mistake but rather the deviation between the measured quantity and the (unknown) true value. For any physical property, there is only one true mean. But if repeated samples are taken, assuming the size of the population is much larger than the size of the sample, differing sample means would be calculated each time. Since all measurements have some amount of error, a single measurement isn't to be trusted. Even taking an average of the sample is an inadequate description of a physical property without considering the uncertainty of that average. Which is why it is necessary to consider the confidence interval of the average. Since 192.607(c)(1) requires performing measurements at 5 places in 2 circumferential quadrants of the pipe. Therefore it is necessary and beneficial to use the measurements to calculate not just the mean value but also the standard deviation. With only 10 measurements, a single outlier can exert significant leverage on the average. In addition to measurement error, the steel itself is not perfectly homogeneous. A variation of at least 3 - 4 ksi in the measured yield strength (YS) around the circumference or end to end is common even when using destructive testing. This variation is in addition to the error introduced by the measurement device itself.

As an example, assume 10 replicate measurements of the YS are taken with a mean of 40 ksi and a standard deviation of 3.5 ksi, on a pipe having an assumed specified minimum YS of 42 ksi. Without accounting for the uncertainty, one might conclude that that the YS is less than the required minimum, which may be an erroneous decision (possible Type I error). The way to account for the uncertainty in the mean is with the confidence interval.

$$
CI = \mu\pm t\frac{s}{\sqrt{n}}
$$ {#eq-confidence}

where: $\mu$ = the average of the sample, s = the sample standard deviation, *n* = the number of measurements, *t* = t-value based on a probability of 0.975, and $n-1$ degrees of freedom (t = 2.26)[^1]

[^1]: The 95% confidence interval refers to the middle 95% of the distribution from 2.5% to 97.5% and the degrees of freedom is the number of samples minus one, (n - 1). The t-distribution is used in statistics to estimate the population parameters for small sample sizes. The formula for the t-distribution is shown in the appendix and the t-value is almost never calculated manually, it is either looked up in statistical tables or more commonly calculated using built-in formulas in a spreadsheet or statistical software.

The 95% confidence interval of the mean for this example, working in units of ksi, would be:

$$
40 \pm 2.26 \left(\frac{3.5}{\sqrt{10}}\right) = 40 \pm 2.50
$$ {#eq-exampleCI}

Conclusion: Since the confidence interval of the mean extends from 37.5 to 42.5 ksi a true yield strength of 42 ksi cannot be excluded and is not rejected. This test could then be counted as being successful for the purposes of verifying the material properties. This is why a measurement that is different should not be treated as inconsistent without considering the uncertainty in the measurement.

An important thing to note about confidence intervals is that it relies on the t-distribution which is dependent on the sample size. As the size of the sample grows the width of the distribution shrinks. There are circumstances where a large sample size is taken and the confidence interval might exclude the reference value and the test would reject the hypothesized value. However, it is important to look at the confidence interval and decide if the difference is a practical difference. For example, if you are trying to determine if the sampled wall thickness could have came from a pipeline with a nominal wall thickness of 0.188 inches and the confidence interval extends from 0.189 to 0.192. While yes, it is true that it is unlikely the mean wall thickness is 0.188, the decision maker needs to decide if that difference is a practical concern before rejecting that sample as inconsistent. As the sample size grows the likelihood of a false negative grows since it will find very small differences that are not of any practical concern. That is why the t-distribution is ideal for sample sizes less than 30. For samples larger than that it is suggested that the critical t value in @eq-confidence be replaced by the critical z value from a normal distribution.

## What Confidence Intervals Tell Us

A common misconception in the interpretation of the confidence interval is that a 95% confidence interval indicates that there is a 95% probability that the true value is within those bounds. The actual meaning of the confidence interval is that if somebody took repeated samples of a given size and calculated the confidence interval each time, 95% of the confidence intervals would include the true value. This is demonstrated in @fig-ci_samp where 100 samples, each with a size of 10 were taken from a population with a known mean of 42. For each sample, the mean and confidence interval is then calculated. Because this plot is based on simulated data, we know the mean is 42, yet exactly half of the sample means are less than the true mean, half are larger and 95 of the 100 confidence intervals include the true value. In practice, someone is not going to take 100 different samples, they are probably going to take one. So the confidence interval is not a probability statement about the mean or the population but a statement about the precision of our estimate of the mean.

This demonstrates why simply using the sample mean is a poor predictor of the population mean. If the sample mean is used by itself to judge the minimum (or maximum) value of the population mean, you are guaranteed 50% of the time the decision will be wrong. The Central Limit Theorem guarantees that near identical results to these (proportion above and below mean and percentage of confidence intervals) will occur regardless of the distribution of the original data, even if it is highly skewed.

```{r}
#| label: fig-ci_samp
#| fig-cap: "Confidence Intervals"
#| dpi: 300
#| fig-width: 6.5


set.seed(1121)
mys <- 42
sdys <- 3.5
pop <- tibble(s = rnorm(1e4, mean = mys, sd = sdys))

samps <-
    rep_slice_sample(
        .data = pop,
        replace = TRUE,
        reps = 100,
        n = 10
    ) %>%
    summarise(
        ms = mean(s),
        lci = t.test(s, mu = mys)$conf.int[1],
        uci = t.test(s, mu = mys)$conf.int[2],
        yn = ifelse(between(mys, lci, uci), "Yes", "No")
    ) %>%
    mutate(id = 1:n())

# samps %>% count(ms>42)

samps %>%
    mutate(yn = factor(yn, levels = c("Yes", "No"))) %>%
    ggplot(aes(id, ms, col = yn)) +
    geom_errorbar(
        aes(
            ymin = lci,
            ymax = uci
        ),
        position = "dodge",
        width = 1,
        linewidth = 0.5
    ) +
    geom_point(aes(y = ms), col = "blue", size = 1) +
    scale_color_manual(values = c("grey70", "red")) +
    geom_hline(
        yintercept = mys,
        lty = 2,
        linewidth = 0.7
    ) +
    theme_bw(14) +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom"
    ) +
    labs(
        x = NULL,
        col = "Includes True Mean",
        y = "Yield Strength (ksi)"
    )

```

The important thing to takeaway from @fig-ci_samp is that the confidence interval is not probability statement about the mean or the population but a statement about the uncertainty in the estimated parameter.

## Comparing Two Samples {#sec-comparing-two-samples}

The previous example is a single sample compared to a reference value (one-sample test). In the alternate scenario where two samples are being compared to see if they are consistent with each other, a similar process is employed. The goal is not to determine consistency with any specific value but rather if it is a reasonable that the two samples could have came from a single population with a common mean. There are several different variations of two-sample t-tests but the relevant one to material verification would be an independent two-sample t-test. In this circumstance, it is assumed that each sample has no influence on the other (independence), the variances are equal and both samples are the same size.

The formula for a independent two-sample t-test is as follows:

$$
t = \frac{\bar{X_1}- \bar{X_2}}{s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}}
$$ {#eq-ttest}

$$
\begin{align*}
&Where: \\
&s_p =Pooled\,standard\,deviation =\sqrt{\frac{s_{X_1}^2 + s_{X_2}^2}{2}}\\
&\bar{X}_{1,2}= Mean\,each\,sample\\
&n_{1,2} = Size\, of\, each\, sample\\
&s_{X_{1,2}} = Standard\,deviation\,of\,each\,sample
\end{align*}
$$

In this case, the the Null is that the means of the two samples are equal. The Null is always based on the assumption that the hypothesis is true and it remains so until the data shows it to be very improbable. The value arrived at in @eq-ttest is the t-statistic which is compared against the critical t-value at a probability of $1-\alpha/2$ with $\nu = n_1 + n_2 - 2$ degrees of freedom. For an $\alpha = 0.05$ with 10 samples in each group, $\nu = 18$, the critical t-value is 2.10. Then if $|t| > t_{1-\alpha/2, \nu}$ then the Null ($H_0$) is rejected in favor of the alternative hypothesis $H_A$. The critical t-value can be calculated from the t-distribution shown in @eq-tpdf (see Appendix) or more commonly looked up in a table or calculated with built-in formulas in any spreadsheet or statistical software[^2]. This test determines if it is plausible based on the samples that one could have come from a population with a different mean than the the other.

[^2]: Example formulas would be, Excel: *T.INV(0.975,18)* or in R: *qt(0.975, 18)*

**Example:**

-   Two samples of size 10 were taken from two populations
-   Pooled standard deviation, $s_p = 5$
-   $\bar{X_1} = 40$
-   $\bar{X_2} = 44$
-   $t = \frac{40-44}{5\sqrt{\frac{1}{10}+\frac{1}{10}}} = -1.79$
-   $t_{critical} = 2.10$

Conclusion: Since $|t| < t_{critical}$ we would not reject that these two samples have a different mean.

In the example of comparing two samples the size of each was given. In the design of a sampling plan the object is to determine the sample size which we know from @sec-sample-size is dependent on the power ($1-\beta$) and acceptable Type I error ($\alpha$). How the power is calculated is dependent on type of data that is being analyzed and all circumstances won't be covered in this paper. There are several online resources and texts that discuss it in detail. The most common case involving two normal distributions like $H_0$ and $H_A$ shown in @fig-type1and2 will be demonstrated here.

## Normal Distribution Sample Size Calculation {#sec-normal-distribution-sample}

In this example, the goal is to be able to detect a difference of more than $\pm$ 3 ksi at a 95% confidence level and a power of 80% between samples from two different pipes. Based on industry data, it is known that the typical standard deviation for the YS is 6 ksi.

Since the comparison is one sample against another, this is a two-sample test. As a simplification to the problem, the normal distribution will be used rather than the t-distribution. The t-distribution is required when the standard deviation is unknown or can't be estimated. But recall that the critical t-value is dependent on the sample size which we are trying to estimate - creating a circular dependency, a "Chicken and Egg" problem where you need the sample size to calculate the t-value but you don't know the sample size you need. To calculate the sample size using a t-test requires an iterative process of guessing a sample size and calculating the power and adjusting the sample size based on the calculated power. For even moderately large samples (generally n \> 30) the difference between the t and normal distribution is small and diminishes as the sample size gets larger. Therefore, since the standard deviation can be estimated, we will use the normal distribution.

The first step is to calculate the critical z-values at $1-\alpha/2$ and $1-\beta$. These values can be looked up in tables or more commonly calculated with the built in functions in any spreadsheet or statistical software.[^3] which in this example, the critical z-values are 1.96 and 0.84 respectively. The critical z-value is the value for of the standard normal distribution at a probability of $1-\alpha/2$ or $1-\beta$.

[^3]: Excel: *NORM.S.INV(0.975) and NORM.S.INV(0.8)*

Then the minimum sample size for a two-sided test is as follows.

$$
N = \frac{2(Z_{1-\alpha/2} + Z_{1-\beta})^2 \sigma^2}{\delta^2}
$$ {#eq-twosample_normal}

$$
\begin{align*}
&Where:\\
&\sigma = Standard\,Deviation\,(6\,ksi)\\
&\delta = Minimum\,difference\,between\,H_0\,and\,H_A\,that\, you\,want\,\,to\,detect\,(3\,ksi)\\
&Z = Critical\,Z\,value\,for\,normal\,distribution\\
&\alpha = Type\,1\,error\\
&\beta = Type\,2\, error
\end{align*}
$$

Putting this all together: $$
N = \frac{2(1.96 + 0.84)^2\times 6.0^2}{3^2} \approx 62.7
$$ Since a fractional sample has no physical meaning, the number is rounded up to the next whole number of 63. If the goal is only a one-sided test the same formula is used with the exception of $Z_{1-\alpha/2}$ is replaced with $Z_{1-\alpha}$. In this example the relatively large required sample size is because of the $\sigma$ to $\delta$ ratio, this scenario is trying to detect a difference of one-half a standard deviation of the population.

In the case of one-sample test where the comparison is a single sample against a reference value the formula is similar to @eq-twosample_normal but with slight differences. If you rearrange the terms of @eq-onesample_normal you will see that it is exactly one-half of @eq-twosample_normal.

$$
N = \left(\frac{Z_{1-\alpha/2} + Z_{1-\beta} }{\delta/\sigma}\right)^2
$$ {#eq-onesample_normal}

## Expanded Sampling Requirements {#sec-expanded-sampling}

As part of this testing requirement under 192.607(e)(4), if a sample is found to be inconsistent with available information or existing expectations or assumed properties used for operations and maintenance in the past, *"the operator must establish an expanded sampling program. The expanded sampling program must use valid statistical bases designed to achieve at least a 95% confidence level that material properties used in the operation and maintenance of the pipeline are valid."* The regulation does not define what is a "statistical valid bases" or how to get to a 95% confidence level. This section will discuss the meaning of confidence level and how it applies to this regulation.

Confidence level is not a probability about the hypothesis being true but rather a statement of how likely we are to make the correct decision based on the testing. In essence, PHMSA is asking the operator to prove to a 95% certainty that the inconsistent material does not exist in the rest of the population. It is a common saying that you can't prove a negative, which is true. However, if an estimate is made (or assumed) about the population it can be shown what the probability of observing certain results are if the estimate was true. Therefore, it can be shown that something is highly improbable based on results, but the negative still can't be proven absolutely.

## Statistically Valid Bases {#sec-statistically-valid-bases}

Significant background has been given up to this point on how to judge a sample against a reference value. As stated previously, the regulation does not define what inconsistent is or how different it needs to be before it's "inconsistent". Therefore it is up to the operator to define this ahead of time. A suggested definition would be if a one-tail test rejects that the sample meets the design threshold.

In an expanded sampling program, each verification dig is the equivalent to a pass/fail test for whether inconsistent results are found or not. This type of sample where the criterion is a binary choice, is known as a Bernoulli trial and is modeled with the binomial distribution. This distribution will determine a probability of *k* successes in *n* independent trials if the probability of success per trial is *p*. A "success" is whatever is being counted. A "success" is not necessarily a positive outcome. In the expanded sampling context, a "success" would be finding an inconsistent material property in an verification dig. The binomial equation is shown as:

$$
\binom{n}{k} p^{k} (1 - p)^{n - k}
$$ {#eq-binom}

Where the $\binom{n}{k}$ is what is called the binomial coefficient which calculates the number of ways *k* successes in *n* trials can be distributed. In an expanded sampling program, the desired number of successes (inconsistent results) is zero but the number of samples required to achieve a 95% certainty is not known a priori. It would be straight forward to show the probability of observing zero successes in a number of trials, but this doesn't show how likely it would be to find something if additional samples were taken.

For instance, the probability of zero successes in one trial is 95% if the inconsistent material is 5% of the population. But even the casual observer would realize that doesn't prove anything, and the result of that one trial is likely the result of random chance and it doesn't demonstrate much about the population. That's equivalent to saying 95% of all cars are white because the first one you saw was white. To show the probability of not finding something in future trials it is necessary to calculate the opposite, the probability of finding one or more successes in *n* trials as shown in @fig-cumprob. Since the total probability of all possible outcomes has to sum to one for a given sample size, the probability of finding one or more is, one minus the probability of finding zero.

Since the desired outcome is not finding any inconsistent samples ( $k=0$ ), the binomial coefficient $\binom{n}{k}$ and $p^k$ in @eq-binom are both equal to 1. So the binomial distribution reduces to simply $(1-p)^n$ where it can be algebraically solved for n directly or iterated in a spreadsheet until $1-(1-p)^n = 0.95$. Where $p$ is the upper limit of the probability of success you are trying to demonstrate (e.g., 0.05).

```{r}
#| label: fig-cumprob
#| fig-cap: "Cumulative Probability"
#| fig-width: 6.5
#| fig-height: 4
#| dpi: 300

size <- 5:70
prob <- c(0.03, 0.05, 0.10, 0.15)


crossed <- expand_grid(size, prob) %>%
  mutate(p_s = pbinom(
    q = 0,
    size = size,
    prob = prob,
    lower.tail = FALSE
  ))

min_samp <- crossed %>%
  filter(p_s >= 0.95) %>%
  group_by(prob) %>%
  summarise(mins = min(size)) %>%
  mutate(p = pbinom(
    q = 0,
    size = mins,
    prob = prob,
    lower.tail = FALSE
  ))


crossed %>%
  ggplot(aes(size, p_s)) +
  geom_line(aes(col = factor(prob)), lwd = 0.9) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(),
    labels = scales::label_percent(),
    limits = c(0.1, 1),
    expand = c(0,0)
  ) +
  scale_x_continuous(breaks = seq(0, 75, by = 10)) +
  geom_hline(
    lty = 2,
    col = 'black',
    yintercept = 0.95,
    lwd = 0.8
  ) +
  geom_segment(data = min_samp,
               aes(
                 x = mins,
                 xend = mins,
                 y = 0.1,
                 yend = p,
                 col = factor(prob)
               ),
               lty = 2,
               lwd=0.8
               ) +
  geom_text(data = min_samp,
            aes(
              x = mins,
              y = 0.25,
              label = paste("P =", prob, "\nN =", mins)
            ),
            size = 5,
            angle =-90,

            ) +
  theme_bw() +
  labs(
    title = "Number of Digs to Confirm 95% Certainty",
    color = "Population Percent",
    y = "Probability of One or More",
    x = "Number of Digs")
```

What @fig-cumprob shows is that the smaller the proportion of the population that is assumed to be inconsistent the greater the number of digs required to reach a 95% confidence level that the inconsistent material doesn't exist elsewhere.

To calculate the necessary sample size, it is necessary to estimate or assume the percent of the population that might be represented by the inconsistent sample. This percentage becomes the probability of success in @eq-binom and the number of successes is set to zero. Then the sample size is iterated until one minus the probability is greater or equal to 95%[^4]. The reason for this is that the proportion of the population affects the probability of finding it in future samples. The larger the proportion it represents, the easier it would be to find in fewer trials and likewise, the smaller the proportion, the more trials it would take to find it even once. What @fig-cumprob illustrates is that if no inconsistent samples are found by the time the probability reaches 95%, then the likelihood of finding additional ones in future digs is 5% or less based on estimated population proportion.

[^4]: The binomial distribution is built into Excel, using that to do the iteration will make this process easier and repeatable.

## Rule of 3 Approximation

A close approximation for sample size is the rule of three. The Rule of Three (also called the "rule of 3") is a simple, widely used statistical rule of thumb for estimating an upper bound on the rate of a rare event (or "defect," "failure," or "inconsistent result") when you observe zero occurrences in a sample of size n. If you test n independent items (or trials) and find zero events of interest, then you can be approximately 95% confident that the true rate (proportion) of that event in the population is less than 3/n. Only applies when you observe exactly zero events. If you see even one, you need a different method (e.g., exact binomial confidence interval).

# Alternate Sampling Plan {#sec-alternate-sampling-plan}

This section presents an alternate sampling plan consistent with 192.607(e)(5), focused on achieving a 95% confidence level using a statistically valid basis. Note that although( e)(5) allows alternative sampling methods, notification to PHMSA for alternate/expanded plans is required under § 192.18. See the regulations for the specifics of notification.

The method presented here uses sequential NDE results to update confidence in a material property (e.g., yield strength). Using Bayes’ Theorem, high confidence can often be achieved with far fewer excavations than the prescriptive approach.).

## Bayes' Theorem

Intuitively, people know that as they acquire more samples that agree with their hypothesis, the probability of it being true should increase as well and it shouldn't take 150 samples. That intuition is correct and the way to show this increase in belief is through Bayes' Theorem.

The goal of an alternate sampling program is to estimate the probability that a hypothesis is true after observing data. Bayes’ Theorem is a natural tool for this (historically referred to as the “inverse probability” problem).

In traditional Frequentist statistics, the hypothesis is specified first and the likelihood of the observed data is evaluated under that hypothesis. In Bayesian statistics, the data are observed and beliefs about the hypothesis are updated based on how well the hypothesis explains the data.

Frequentist methods often require larger samples because they intentionally avoid incorporating prior knowledge into the calculation and instead rely on long-run sampling behavior. Bayes’ Theorem works well when data are sparse but meaningful prior information is available.

Bayesian statistics starts with a prior distribution (“the prior”) and updates it as new information is acquired. The prior is the starting point, and it is conditioned (updated) in proportion to the weight of new evidence.

### Nomenclature {#sec-nomenclature}

In the following example when the notation of P followed by parentheses such as P(*x*) this represents the probability of the independent event happening, the probability of observing *x* in this case. When there are two variables separated by a vertical bar such as P(H\|*x*) this is a conditional probability. In this example it would be the probability of the hypothesis, H given that *x* (the data) has been observed. It is read as the probability of H given *x*.

The most basic form of Bayes Theorem is represented by the following the equation.

$$
P(A|B) = \frac{P(B|A)P(A)}{P(B)}
$$ {#eq-bayes}

There are two main components to Bayes' Theorem, the hypothesis (the criterion of interest) and the evidence (data). When there is only a binary condition being considered (e.g., true/false, sick/not sick) Bayes' Theorem can be written conceptually as:

$$
P(H|E)=\frac{P(E|H)P(H)}{P(E)}=\frac{P(E|H)P(H)}{P(E|H)P(H)+P(E|H')P(H')}
$$ {#eq-bayes_phe}

Where:

P(H\|E) = Probability of hypothesis is true given the evidence (e.g. probability of disease given a positive test)

P(H) = Probability that hypothesis is true overall (e.g. percentage of population that actually have the disease)

P(E) = Probability of observing the evidence if the hypothesis is true. In the case of disease diagnosis, it would be the P(positive test if disease is true)+P(positive test if no disease), i.e. True positive rate + False positive rate

P(E\|H) = Conditional probability of the observing the evidence if the hypothesis is true. (e.g. probability of a positive test if the disease is present)

P(E\|H') = Conditional probability of observing the evidence if the hypothesis is false (e.g. probability of positive test if the disease is not present)

P(H') = Probability the hypothesis is false (e.g. probability the disease is not present)

In the above construct, H' represents the hypothesis being false.

## Sequential Updating

The method for sequential updating of a Bayesian inference starts with the prior probability (prior) of the event happening and a probability of the test correctly identifying the value of interest. Using this prior information and known test accuracy an updated probability is calculated after each round of testing. Then after each test the updated posterior becomes the prior for the next test. This is illustrated with the following example. The prior is simply a statement of the strength user's belief that the hypothesis is true based on what is known a priori about the pipeline. This can be based on design standards and operating pressure. It's not intended to be a long-running frequency but rather how confident the user is in the hypothesis based on any prior information before seeing the data.

The average person uses prior probability all the time without consciously thinking about it. If someone makes an extraordinary claim, you are skeptical. Why? Because you placed such a small prior probability on it being true based on your experience and knowledge. This is congruent with a principle established by the mathematician Pierre-Simon Laplace, "The weight of evidence for an extraordinary claim must be proportioned to its strangeness"[^5]. Even if you were to learn that the person making the claim is truthful 90% of the time you would still remain unconvinced. This is because even the 90% rate of truthfulness isn't enough to overcome the very small prior you placed on the claim. You didn't consult a database of such claims to develop the prior, you knew from general knowledge and experience that the probability of truthfulness of that claim was very low. Your prior is just the starting point, as data is accumulated it will move the updated probability either toward or away from the prior.

[^5]: Pierre-Simon Laplace, *A Philosophical Essay on Probabilities*, trans. Frederick Wilson Truscott and Frederick Lincoln Emory (New York: John Wiley & Sons, 1902), 17.

In this example, nondestructive testing is used to determine an unknown YS. It is known from qualification testing that the tool will produce results that are either correct or conservative 60% of the time. Based on prior knowledge of the operating pressure and construction practices in the time frame of its installation it is believed that the true YS is 42 ksi or greater. But given that the information about the YS is not documented, the initial belief of that is limited to a skeptical 10% (i.e., a 90% probability that it does not meet the minimums). If the information was more reliable, the decision maker might start at 20 or 30% depending on the level of confidence in the prior information they have. The initial belief level should be dependent on the amount and quality of the prior information that is available before testing. It's the starting point not the final destination.

Assuming the first test is successful (YS $\geq$ 42), the probability of being YS $\geq$ 42 ksi ($YS_{42}$) is updated according to @eq-bayes as follows.

$$
P(YS_{42}) = \frac{(0.10)(0.60)}{(0.10)(0.60)+(1-0.10)(1-0.60)}=14.3\%
$$ {#eq-examp1}

The 14.3% then becomes the prior for the next test. If the second test is successful then the updated probability becomes

$$
P(YS_{42}) = \frac{(0.143)(0.60)}{(0.143)(0.60)+(1-0.143)(1-0.60)}=20.0\%
$$ {#eq-examp2}

Then each test is updated in the same way until the probability of success meets or exceeds 95% as shown in @fig-update. In this example, it is achieved after 13 samples. This example assumes that the population is homogeneous and any variability is due to the uncertainty in the measurement. The actual number of confirming samples required will depend on the prior probability used and the accuracy of the NDE method used.

The caveat of this calculation is it is only attempting to infer if the pipeline meets some threshold value. We are not concerned about what the true value is, only the binary hypothesis that it meets a minimum threshold value or not.

```{r}
#| label: fig-update
#| fig-cap: "Updated Beliefs"
#| fig-width: 6.5
#| fig-asp: 0.67
#| dpi: 300


pS <- 0.10
pT <- 0.60

p0 <- pS*pT/(pS*pT+(1-pS)*(1-pT))

n=0:15

p2=pS*pT^n/(pS*pT^n+(1-pS)*(1-pT)^n)

updated <- tibble(n,p2)

ggplot(updated, aes(n, p2)) +
  geom_point(col = 'blue', size = 3) +
  geom_line(col = 'blue', lwd = 1.2) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
   # geom_label(aes(
   #   x = n,
   #   y = p2 - .05,
   #   label = round(p2, 2)
   # ),
   # size = 3) +
  theme_bw() +
  scale_x_continuous(breaks = scales::pretty_breaks(),
                     minor_breaks = 1:max(n)) +
  labs(title = "Probability of YS > 42 ksi",
       y = "Probability",
       x = "Number of Tests")
```

What @fig-update shows is that by using Bayes' Theorem and sequential updating, a 95% confidence level can be achieved in less than one-tenth the number of digs required by the prescriptive requirements (up to 150 digs).

This same process could, in theory, be used for expanded sampling. However, the premise here is that the population is homogeneous and you are confirming whether it meets a criterion (a binary threshold). In contrast, the need for expanded sampling suggests the population may not be homogeneous, so it may be difficult to justify this base assumption.

That said, there is a Bayesian framework that can be used to reach a 95% level of certainty with fewer samples than what was shown in @sec-expanded-sampling .

## Bayesian Updating of a Binomial Process

The binomial distribution was explained in @sec-statistically-valid-bases and how it is used to infer the probability of success. In Bayesian statistics there are families of distributions that are known as conjugate priors. Conjugate priors are an algebraic convenience that allow the user to model the underlying process in a Bayesian framework using a simplified closed-form equation. When working with a binomial data generating process, the conjugate prior is Beta distribution. The advantage of the beta distribution is that it is constrained to between 0 and 1 which makes it ideal to modeling proportions.

The shape of the beta distribution is controlled by two shape parameters $\alpha$ and $\beta$, not to be confused with the Type I and II errors which use the same symbols. When referring to the distribution with a given set of parameters it is written as $Beta(\alpha,\, \beta)$. The mean of the distribution is $\frac{\alpha}{\alpha+\beta}$ and variance $\frac{\alpha \beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$. Several examples of different beta distribution priors are shown in @fig-betadist.

```{r}
#| label: fig-betadist
#| fig-cap: "Examples of Beta Distributions"
#| dpi: 300
#| fig-width: 6.5
#| fig-height: 5


# 1. Define parameter combinations
params <- expand.grid(
  alpha = c(0.5, 1, 5), 
  beta  = c(0.5, 1, 5)
)

# 2. Generate x values
x_vals <- seq(0, 1, length.out = 250)

# 3. Create the data frame and format the labels with Greek letters
plot_data <- params %>%
  group_by(alpha, beta) %>%
  reframe(
    x = x_vals,
    y = dbeta(x_vals, alpha, beta),
    # Use plotmath syntax for Greek letters: alpha and beta
    label = paste0("alpha == ", alpha, " ~ ',' ~ beta == ", beta)
  )

# 4. Create the plot
ggplot(plot_data, aes(x = x, y = y)) +
  geom_line(color = "firebrick", size = 1) +
  geom_area(fill = "firebrick", alpha = 0.1) +
  # 'labeller = label_parsed' is the key to rendering Greek letters
  facet_wrap(~label, scales = "free_y", ncol = 3, labeller = label_parsed) +
  labs(
    title = expression(paste("Shapes of the ", Beta, " Distribution")),
    subtitle = "Visualizing the influence of shape parameters",
    x = expression(theta),
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "gray95", color = NA),
    strip.text = element_text(face = "bold")
  )

```

When $\alpha = \beta$ and both are greater than 1 it creates a symmetrical bell-shaped distribution similar to a normal distribution. When $\alpha$ and $\beta = 1$ the distribution is uniform and when one of the parameters is larger than the other it creates a skewed distribution with the direction of the skew dependent on which one is larger.

Here's how a binomial updating process in the context of an expanded sampling program would work. As in any Bayesian process it is necessary to start with a prior. The prior can be thought of a statement of certainty in the population proportion before any observations are completed. A diffuse prior is an indication of large uncertainty in the proportion while a narrow/sharply-peaked indicates that the user is confident on the population proportion. A widely diffuse prior is known as a "non-informative" prior meaning it has minimal influence on the posterior distribution. The suggested prior for a expanded sampling program would be the $Beta(0.5,0.5)$ shown in top left one in @fig-betadist. This is non-informative prior the peaks are at 0 and 1 which means that it will update quickly one way or the other as observations are completed.

Updating the beta distribution is quite simple. After *n* observations with *x* successes your updated distribution becomes $Beta(\alpha + x,\,\beta+n-x)$. For example if after 10 samples with no inconsistent material found, the updated distribution would be $Beta(0.5+0,\, 0.5+10-0) = Beta(0.5,10.5)$ shown in @fig-updatebeta.

```{r}
#| label: fig-updatebeta
#| fig-cap: "Updated Distribution"
#| fit-width: 5
#| fig-height: 3.5
#| dpi: 300

pgt <- 1- pbeta(0.05, 0.5,10.5)
tibble(x = seq(0.001, 0.25, length.out = 301), y = dbeta(x, 0.5, 10.5)) |>
  ggplot(aes(x, y)) +
  geom_line(col = "firebrick", size = 1) +
  geom_area(fill = "firebrick", alpha = 0.1) +
  geom_vline(xintercept = 0.05, lty = 2, size = 1) +
  annotate("text", x = 0.125, y = 20, label = latex2exp::TeX("$P(\\theta > 0.05) = 0.31$"), size =5)+
  theme_minimal() +
  labs(
    y = "Density",
    x = expression(theta),
    title = "Updated Beta Distribution"
  )
  
```

The distribution of credible values is heavily concentrated on the left but there is still a significant tail to the right. To determine the probability that the proportion of inconsistent material is greater than 5% the cumulative distribution function (CDF) is used. The CDF is simply the integral of the probability density function (PDF) from 0 to some value. In this case we want to know the probability that the proportion is greater than 5% which is one minus the CDF at 5%. This is calculated as $1-P(\theta \leq 0.05)$ where P is the CDF. Using statistical software or a spreadsheet this can be calculated easily. In this case it comes out to 31% meaning there is a 31% chance that the proportion of inconsistent material is greater than 5% even after 10 samples with no inconsistent reContinuing this process until the total probability of success drops below 5% gives the number of samples required to demonstrate (with 95% certainty) that the proportion of inconsistent material is less than a given percentage. In this example, as the number of digs increases with no inconsistent samples found, the distribution becomes increasingly narrow and concentrated to the left. the left.

```{r}
#| label: fig-betacum
#| fig-cap: "Cumulative Probability from Beta Distribution"
#| dpi: 300
#| fig-width: 6
#| fig-height: 4.5

# Define the prior parameters (Alpha=0.5, Beta=0.5 is Jeffreys Prior)
alpha_prior <- 0.5
beta_prior <- 0.5

# Define the scenarios: 0 successes and varying numbers of failures
successes <- 0
failures_vec <- c(0, 10, 20, 30, 40) # 0 failures represents the prior itself

# Create a sequence of probability values (x-axis)
theta <- seq(0, 1, length.out = 500)

# Generate the data for each scenario
plot_data <- data.frame(theta = theta)

for (f in failures_vec) {
  # Posterior parameters: alpha = alpha + successes, beta = beta + failures
  a_post <- alpha_prior + successes
  b_post <- beta_prior + f
  
  label <- paste0("Failures: ", f, " (Beta ", a_post, ",", b_post, ")")
  plot_data[[label]] <- dbeta(theta, a_post, b_post)
}

# Reshape data for ggplot
plot_long <- pivot_longer(plot_data, cols = -theta, names_to = "Scenario", values_to = "Density")

# Create the plot
ggplot(plot_long, aes(x = theta, y = Density, color = Scenario)) +
  geom_line(size = 1) +
  ggsci::scale_color_jco() +
  labs(
    title = "Beta-Binomial Updating (0 Inconsistent, Increasing No. of Digs)",
    subtitle = paste0("Prior: Beta(", alpha_prior, ", ", beta_prior, ")"),
    x = expression(theta ~ "(Probability of Success)"),
    y = "Density"
  ) +
  geom_vline(xintercept = 0.05, lty =2, alpha = 0.5)+
  theme_minimal() +
  coord_cartesian(ylim = c(0, 10), xlim = c(0,0.25),expand = c(0,0)) # Zoom in to see the shape clearly


```

# Conclusion {#sec-conclusion}

The methodologies presented in this paper provide a robust and defensible framework for meeting the material verification requirements of 49 CFR 192.607 by transitioning from prescriptive sampling to statistically valid alternate plans, operators can achieve the required 95% confidence level with significantly greater efficiency.

Successful implementation relies on three key pillars:

-   **Risk-Based Decision Making**: The selection of error thresholds must be driven by the specific consequences of Type I versus Type II errors. Because these risks are intertwined, a sampling plan must explicitly define which error is of greater concern and set thresholds accordingly.

-   **Measurement Integrity**: Operators must account for both measurement uncertainty and the inherent non-homogeneity of steel. Utilizing Confidence Intervals rather than simple sample means ensures that a single outlier does not trigger an unnecessary and costly "expanded sampling" program.

-   **Optimization via Bayesian Logic**: For populations lacking TVC records, the application of Bayes’ Theorem offers a statistically sound path to verify properties like yield strength using a fraction of the 150 excavations typically required. This approach allows for the incorporation of prior knowledge, such as vintage-specific manufacturing standards, to reach a 95% belief level through sequential NDE testing.

Ultimately, these principles empower the end-user to define a sampling program that is not only regulatory compliant but also economically sustainable. By grounding the plan in first principles—from Bernoulli trials for expanded sampling to t-distributions for mean comparisons—operators can confidently explain their process to regulators and stakeholders alike.

# Appendix A: Mathematical Formulas

## t-distribution formula

The probability density function (pdf) of the t-distribution with $\nu$ degrees of freedom is given by:

$$ 
f(t|\nu) = \frac{\Gamma(\frac{\nu+1}{2})}{\sqrt{\nu \pi}\Gamma(\frac{\nu}{2})}\left(1+\frac{t^2}{\nu}\right)^{-\frac{\nu+1}{2}}
$$ {#eq-tpdf}

Where $\Gamma$ is the gamma function:\
$$ 
\Gamma(z) = \int_{0}^{\infty} x^{z-1} e^{-x} dx
$$ {#eq-gamma}

This is provided mostly for information purposes. The critical t-values and probabilities can be calculated with the built-in functions in almost any spreadsheet or statistical software.

## Z- distribution

$$
f(z) = \frac{1}{\sqrt{2\pi}} e^{-\frac{z^2}{2}}
$$

where: $$
z = \frac{\bar{x} - \mu}{\sigma}
$$ For two-tailed 95% confidence ($\alpha = 0.05$): $$
z_{\alpha/2} = z_{0.025} \approx 1.96
$$

For one-tailed 95% confidence (or 5% significance in one direction): $$
z_{\alpha} = z_{0.05} \approx 1.645
$$

# Appendix B Definitions for Material Verification {#sec-definitions}

This section aims to bridge the gap between abstract statistical theory and the operational decision-making required by 49 CFR 192.607. These definitions are intended to provide a concise reference of the statistical terms used in this paper.

## Null Hypothesis

The null hypothesis is a statement of no effect or no difference. In the context of material verification, the null hypothesis is that the material meets the specified requirements. The statistical test is designed to determine if there is sufficient evidence to reject the null hypothesis. This process is known as a Null Hypothesis Significance Test (NHST).

## Inconsistent Result

A material test result is defined as inconsistent if it meets one of the following criteria:

-   **Statistical Exclusion**: The hypothesized or documented value (e.g., SMYS or Nominal Wall Thickness) falls outside the 95% Confidence Interval calculated from the sample measurements.

-   **Practical Significance**: Even if statistically different, a result is only "inconsistent" if the deviation is of a practical concern to the safe operation or maintenance of the pipeline.

## Population (Lot)

In alignment with 192.607(e)(1), a population is a grouping of pipeline segments—not necessarily contiguous—defined by a unique combination of:

-   **Material Attributes**: Grade, nominal wall thickness, and manufacturing process.

-   **Vintage Data**: Pipe manufacturing dates and construction dates.

-   **Statistical Independence**: For the purposes of sampling, the lot size should be at least 10 times greater than the sample size to ensure the sample remains representative without requiring lot-dependent size adjustments.

## Acceptance Thresholds (Error Rates)

The following thresholds are recommended to balance safety with economic sustainability:

-   **Type I Error (**$\alpha$): Typically set at 0.05 (5%). This limits the probability of "false positives"—incorrectly rejecting valid material records and triggering unnecessary excavations.

-   **Type II Error (**$\beta$): Typically set at 0.20 (20%), targeting a test power of 80%. This ensures a high probability of correctly identifying truly substandard material while maintaining a realistic sample size.

These should be considered guidelines, the acceptable Type I and II error is dependent on the application. See @sec-statistical-tests-and-errors for a more detailed discusion.

## Statistically Valid Basis

A statistically valid basis under 192.607(e)(5) is not defined in the regulations, but is generally understood to be one of the following:

-   **Frequentist Hypothesis Testing**: A one-sample or two-sample t-test or Z-test designed to detect a specific Effect Size ($d$).

-   **Bayesian Sequential Updating**: A process that starts with a Prior Probability (based on manufacturing standards or operating history) and updates the belief level with each successful NDE test until a 95% posterior probability is achieved.